---
title: "Linear Models"
output: html_notebook
---

## Introduction

Money ball data is used through out the lecture series.

## Load Data Analysis Libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(dslabs)
```

## Load the data

```{r}
library(Lahman)
```


```{r}
str(Teams)
```

```{r}
head(Teams)
```

## Create variables to analyze the dataset

Lets analyze a few variables against Runs per game
-> Home runs per game
-> Stolen Bases per game
-> Base on Balls per game

Lets compute these variables

```{r}
T2002 <- Teams %>% filter(yearID %in% 2002) %>% 
                      mutate (
                          singles = H - HR - X2B - X3B, 
                          R_per_game = R/G, 
                          HR_per_game = HR/G, 
                          SB_per_game = SB/G, 
                          BB_per_game = BB/G, 
                          AB_per_game = AB/G, 
                          singles_per_game = singles / G, 
                          doubles_per_game = X2B / G, 
                          triples_per_game = X3B / G
                          )

Teams <- Teams %>% filter(yearID %in% 1961:2001) %>% 
                      mutate (
                          singles = H - HR - X2B - X3B, 
                          R_per_game = R/G, 
                          HR_per_game = HR/G, 
                          SB_per_game = SB/G, 
                          BB_per_game = BB/G, 
                          AB_per_game = AB/G, 
                          singles_per_game = singles / G, 
                          doubles_per_game = X2B / G, 
                          triples_per_game = X3B / G
                          )
str(Teams)
```

```{r}
Teams
```

## Polt the variables created

### Plot: home runs per game vs. runs per game

```{r}
ds_theme_set()
Teams %>% ggplot(aes(HR_per_game, R_per_game)) + 
  geom_point(alpha = 0.5) +
#  geom_abline()
  stat_smooth(method = "lm") # Adds a 
```


### Plot: stolen bases per game vs. runs per game

```{r}
Teams %>% ggplot(aes(SB_per_game, R_per_game)) + 
  geom_point(alpha = 0.5) +
#  geom_abline()
  stat_smooth(method = "lm") # Adds a 
```

### Plot: bases on ball per game vs. runs per game

```{r}
Teams %>% ggplot(aes(BB_per_game, R_per_game)) + 
  geom_point(alpha = 0.5) +
#  geom_abline()
  stat_smooth(method = "lm") # Adds a 
```

### Plot: at-bats per game vs. runs per game

```{r}
Teams %>% ggplot(aes(AB_per_game, R_per_game)) + 
  geom_point(alpha = 0.5) +
#  geom_abline()
  stat_smooth(method = "lm") # Adds a 
```

## Confuounding

### Higher BB => Higher Runs Per Game.
However, this is not true if we understand the BaseBall game bette.  Homeruns usually results in BB and this is why BB is higher.


Lets check the correlation between BB & Home runs.  Also compare Singles with Home runs.

```{r}

Teams %>% summarize(
                                                      cor(HR_per_game, BB_per_game), 
                                                      cor(HR_per_game, singles_per_game), 
                                                      cor(BB_per_game, singles_per_game)
                                                    )

```

Bases on Balls are confounded with Home Runs



## lm function


```{r}

fit <- lm(R_per_game ~ HR_per_game + BB_per_game, Teams)
fit
```

```{r}
summary(Teams)
```

## Tibbles

Tibbles display much better

### Differences with Dataframe

### Dataframe display

```{r}
Teams
```

### Tibbles display

```{r}
library(tidyverse)
library(tidyquant)

t <- as_tibble(Teams)
t
```

### DO Function

Create a regression line for each strata

```{r}

dat <- Teams %>% filter(yearID %in% 1961:2001) %>% 
  mutate(HR = round(HR/G, 1), 
         BB = BB / G, 
         R = R / G) %>% 
  select(HR, BB, R) %>% 
  filter(HR >= 0.4 & HR <= 1.2)

dat %>% group_by(HR) %>% do(fit = lm(R ~ BB, data = . ) )

```

Create a function that returns a dataframe instread of a function

```{r}

get_slope <- function(data) {
  fit <- lm(R ~ BB, data = data)
  sum.fit <- summary(fit)

  data.frame(slope = sum.fit$coefficients[2, "Estimate"], 
             se = sum.fit$coefficients[2, "Std. Error"],
             pvalue = sum.fit$coefficients[2, "Pr(>|t|)"])
}

dat %>% group_by(HR) %>% do(get_slope(.))

```

To know whether the relationship between home runs and runs per game varies by baseball league.

```{r}
library(broom)
dat <- Teams %>% select(lgID, HR_per_game, BB_per_game, R_per_game) 

dat %>% 
  group_by(lgID) %>% 
  do(glance(lm(R_per_game ~ HR_per_game, data = .)))

```

## Building a Better Offensive Metric for Baseball

```{r}
fit <- Teams %>% lm(R_per_game ~ HR_per_game + BB_per_game, data = .)

tidy(fit)

```

```{r}
tidy(fit, conf.int = TRUE)
```

```{r}
fit <- Teams %>% lm(R_per_game ~ BB_per_game + singles_per_game + doubles_per_game + triples_per_game + HR_per_game, data = .)
tidy(fit, conf.int = TRUE)
```

Lets predit for 2002 based on the above model

```{r}
t <- T2002 %>% mutate(R_pred = predict(fit, newdata = .))
```

```{r}
t %>% ggplot(aes(R_pred, R_per_game)) + 
  geom_point() + 
  geom_text(aes(label = teamID), nudge_x = 0.06) + 
#  stat_smooth(method = "lm") + 
  geom_abline(slope = 1, intercept = 0)
```

## Modelling individual player

Lets look at batting dataset

```{r}
help(Batting)
```

```{r}
pa_per_game <- Batting %>% filter(yearID == 2012) %>% 
  group_by(teamID) %>% 
  summarise(pa_per_game = sum(AB + BB) / max(G)) %>% 
  .$pa_per_game %>% mean

pa_per_game
```

### Predict

Imagine you have two teams. 
Team A is comprised of batters who, on average, get two bases on balls, four singles, one double, and one home run. 
Team B is comprised of batters who, on average, get one base on balls, six singles, two doubles, and one triple.


#### Create a data frame for this data
```{r}

nd <- data.frame(teamId = c("A", "B"), BB_per_game = c(2, 1), singles_per_game = c(4, 6), doubles_per_game = c(1, 2), triples_per_game = c(0, 1), HR_per_game = c(1, 0))

nd

```

#### Predit the data

```{r}
predict(fit, newdata = nd)
```

To be sure what data corresponds to which column, we can mutate the column into dataframe.

```{r}
nd <- nd %>% mutate(R_pred = predict(fit, newdata = .))
nd
```

