---
title: "Correlation & Stratification"
output: html_notebook
---

## Introduction



## Loading data

Lets load the Golton Families data

```{r}
library(HistData)
library(tidyverse)
data("GaltonFamilies")
str(GaltonFamilies)
```

## Filter Data

Create a subset with
-> Only first sons ( gender = "male") ( childNum = 1)
-> Only Father and child's height

```{r}
 galton_heights <- GaltonFamilies %>% filter(childNum == 1 & gender == "male") %>% select(father, childHeight) %>% rename(son = childHeight)
```

Lets plot and see if there are related

```{r}
library(dslabs)
ds_theme_set()
galton_heights %>% ggplot(aes(father, son)) + 
  geom_point() + 
  stat_smooth(method = "lm")
  
```

## Compute Correlation

Since the data is normal distributed lets find the mean & SD
```{r}
galton_heights %>% summarize(mean(father), sd(father), mean(son), sd(son), cor(father, son))
```

Lets compute correlation through its formula

```{r}
mu_f <- mean(galton_heights$father)
s_f <- sd(galton_heights$father)
mu_s <- mean(galton_heights$son)
s_s <- sd(galton_heights$son)

# Lets compute (1) the father's height distance from mean (2) sons height distance from mean (3) product of these
gh <- galton_heights %>% mutate(df = (father - mu_f) / s_f, ds = (son - mu_s) / s_s ) %>% mutate(corx = df * ds)
# correlation = sum of the product by count of the product
gh %>% summarize(sum(corx)) / count(gh)
```

We could as well rewrite the entire above code in the below one line

```{r}
# Lets compute (1) the father's height distance from mean (2) sons height distance from mean (3) product of these
galton_heights %>% mutate(df = (father - mean(father)) / sd(father), ds = (son - mean(son)) / sd(son) ) %>% mutate(corx = df * ds) %>% summarize(sum(corx)) / count(galton_heights)
```

## Stratification

### Problem: 
Find height of son from fathers height ( father's height = 72 ) 

### 1. Simple prediction 
Take the mean irrespective of father's height

```{r}
mean(galton_heights$son)
```

### 2. Using correlation 
Given father's height is 72 lets figure out how far is he from mean apply the same to Son.

Mean:
```{r}
mean(galton_heights$father)
```

SD: 
```{r}
sd(galton_heights$father)
```
Distance from mean in terms of standard deviation: 
```{r}
dmf = (72 - mean(galton_heights$father)) / sd(galton_heights$father)
dmf
```

Calculating sons height
```{r}
dmf * sd(galton_heights$son) + mean(galton_heights$son)
```
This is an over estimate.

### 3. Stratification

Round off the heights of father to nearest number and take the average from there.

Below chart shows the rounded fathers heights against sons height

```{r}
ds_theme_set()
galton_heights %>% ggplot(aes(round(father), son)) + 
  geom_point() + 
  stat_smooth(method = "lm")
```

```{r}
galton_heights %>% mutate(rf = round(father)) %>% filter(rf == 72) %>% summarize(mean(son))
```

### Plots to view stratified data

#### Box plot to view the distribution of sons height data at rounded of height

```{r}
galton_heights %>% mutate(Father_Height_Factor = factor(round(father))) %>% ggplot(aes(Father_Height_Factor, son)) + 
  geom_boxplot() +
  geom_point() 
```

#### Plot of fathers rounded height vs. sons height mean

```{r}
galton_heights %>% mutate(father = round(father)) %>% group_by(father) %>% summarize(mean(son))
```

```{r}
galton_heights %>% mutate(father = round(father)) %>% group_by(father) %>% summarize(ms = mean(son)) %>%
  ggplot(aes(father, ms)) +
  geom_point() 
```

#### Draw the scaled graph with slope

Find the correlation
```{r}
r = cor(galton_heights$father, galton_heights$son)
r
```

```{r}
require(stats)
galton_heights %>% mutate(father = round(father)) %>% group_by(father) %>% summarize(son = mean(son)) %>%
  mutate(z_father = scale(father), z_son = scale(son)) %>% 
  ggplot(aes(z_father, z_son)) +
  geom_point() + 
  geom_abline(slope = r, intercept = 0)
```
### Understanding scale function 

Scale documentation says: 
"scale is generic function whose default method centers and/or scales the columns of a numeric matrix."

When compared default scale values to distance from the mean its exactly the same.  See below:
```{r}
galton_heights %>% mutate(ss = scale(son), sf = scale(father), zf = (father - mu_f) / sd_f, zs = (son - mu_s) / sd_s )
```


### Add the regression line to the father and son hights chart

Slope = Corr(x, y) * SD (y) / SD (x)
Slope = Corr(Father, Son) * SD(son) / SD(father)

Intercept = mean(y) - slope * mean(x)


```{r}
mu_f <- mean(galton_heights$father)
sd_f <- sd(galton_heights$father)
mu_s <- mean(galton_heights$son)
sd_s <- sd(galton_heights$son)
r <- cor(galton_heights$father, galton_heights$son)
m <- r * s_s / s_f
b <- mu_s - m * mu_f

galton_heights %>% ggplot(aes(father, son)) +
  geom_point() + 
  geom_abline(slope = m, intercept = b)

```

## Bivariate normal distribution

Two variables are said to be in bivariate normal distribution if for each grouping of X, Y is in normal distribution.


```{r}
galton_heights %>% mutate(z_father = round( (father - mu_f) / sd_f ) ) %>%
  filter(z_father %in% -2:2) %>%
  ggplot() + 
  stat_qq(aes(sample = son)) + 
  facet_wrap(~z_father)
```

## Linear Models

```{r}
mod <- lm(son ~ father, data = galton_heights)
summary(mod)
```

## Least Squares Estimates (LSE)

Find least value of Beta1 where Beta0 = 36

```{r}

rss <- function(beta0, beta1, data) {
  resid <- galton_heights$son - (beta0 + beta1 * galton_heights$father)
  return(sum(resid * resid))
}

beta1 = seq(0, 1, len=nrow(galton_heights))
results <- data.frame(beta1 = beta1,
                      rss = sapply(beta1, rss, beta0 = 36))
results %>% ggplot(aes(beta1, rss)) + geom_line() + 
  geom_line(aes(beta1, rss), col=2)
```

### LSE are random variables

```{r}
B <- 1000
N <- 100
lse <- replicate(B, {
  sample_n(galton_heights, N, replace = TRUE) %>% 
    lm(son ~ father, data = .) %>% .$coef 
})

lse <- data.frame(beta_0 = lse[1,], beta_1 = lse[2,]) 

library(gridExtra)

p1 <- lse %>% ggplot(aes(beta_0)) + geom_histogram(binwidth = 5, color = "black")
p2 <- lse %>% ggplot(aes(beta_1)) + geom_histogram(binwidth = 0.1, color = "black")

grid.arrange(p1, p2, ncol = 2)

```

### Plots to predict the sons height

```{r}
galton_heights %>% ggplot(aes(father, son)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
library(tidyquant)

model <- lm(son ~ father, data = galton_heights)
predictions <- predict(model, interval = c("confidence"), level = 0.95)
data <- as.tibble(predictions) %>% bind_cols(father = galton_heights$father)

ggplot(data, aes(x = father, y = fit)) +
  geom_line(color = "blue", size = 1) + 
  geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=0.2) + 
  geom_point(data = galton_heights, aes(x = father, y = son))
```





