---
title: "Tidy Data"
output: html_notebook
---

This notebook is student work of Tidy Data chapter as explained in the "Data Science: Wrangling" course.


Load gapminder module

```{r}
library(gapminder)
library(dplyr)
data(gapminder)

```

Read gapminder data and filter it

```{r}
head(gapminder)
```

```{r}
tidy_data <- gapminder %>% filter(country %in% c("South Korea", "Germany", "India", "Singapore") ) %>% select(country, year, lifeExp)
```

Plot the data

```{r}
library(ggplot2)
tidy_data %>% ggplot(aes(year, lifeExp, col = country)) + geom_point()
```

<H3> Gather - Transpose wide data into column formmat </H3>

Lets get the wide formatted data from dslabs

```{r}
library(readr)
path <- system.file("extdata", package = "dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
head(wide_data)
```



Reshape the data

```{r}
library(tidyverse)
new_tidy_data <- wide_data %>% gather(year, fertility, '1960':'2015')
# Can also be rewritten as 
new_tidy_data <- wide_data %>% gather(year, fertility, -country)
head(new_tidy_data)
```

Convert the type to integer

```{r}
new_tidy_data <- wide_data %>% gather(year, fertility, -country, convert = TRUE)
head(new_tidy_data)
```

<H3> Spread - Transpose tidy data into wide data </H3>

New spread the fertility data by country

```{r}
new_wide_data <- new_tidy_data %>% spread (country, fertility)
head(new_wide_data)
```

<H3> Separate and unite the data</H3>

Read the data


```{r}
path <- system.file("extdata", package = "dslabs")
filename <- file.path(path, "life-expectancy-and-fertility-two-countries-example.csv")
raw_data <- read_csv(filename)
head(raw_data)
```

The problem is there are two variables life_expectancy and fertility appended to year.

So we will run: 
1) Create a column VAR with values ( 1960_fertility, 1960_life_expectancy .... )
2) Split the column into 2 columns, one having year and other the variable name ( life expectancy or fertility)
3) Use the spread function to create 2 columns life_expectancy & fertility

Step 1: 
```{r}
s1_data <- raw_data %>% gather(var, val, -country)
head(s1_data)
```

Step 2: 
```{r}
s2_data <- s1_data %>% separate(var, c("year", "variable_name"), sep = "_", extra = "merge" )
head(s2_data)
```

Step3: 
```{r}
s3_data <- s2_data %>% spread( variable_name, val )
head(s3_data)
```

<H3> Combining Tables </H3>

Review the datasets used for exercise

```{r}
library(dslabs)
head(murders)
```


```{r}
head(results_us_election_2016)
```


Join murders and results_us_election_2016 datasets 


```{r}
joined_ds <- left_join(murders, results_us_election_2016, by = "state") %>% select(state, abb, region, population, electoral_votes)
head(joined_ds)
```

Take subset of data from these 2 tables to run various joins

```{r}
tab1 <- slice(murders, 1:6) %>% select(state, population)
tab2 <- slice(results_us_election_2016, c(1:3, 5, 7:8) ) %>% select(state, electoral_votes)
```

<b>Inner join</b> - joins two datasets by the key and only returns matching datasets

```{r}
inner_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% inner_join(tab2, by = "state")
```


<b>Left join</b> - results in taking everything from left dataset ( tab1 ) and matching data from tab2

```{r}
left_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% left_join(tab2, by = "state")
```


<b>Right join</b> - results in taking in everything from right dataset ( tab2 ) and matching data from tab1

```{r}
right_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% right_join(tab2, by = "state")
```


<b>Full join</b> - results in taking in everything from left dataset ( tab1 ) right dataset ( tab2 )

```{r}
full_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% full_join(tab2, by = "state")
```


<b>Semi join</b> - keeps the records in left dataset ( tab1 ) if there are matching records in right dataset ( tab2 )
Semi Join will not bring any columns from right dataset ( tab 2 )

```{r}
semi_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% semi_join(tab2, by = "state")
```

<b>Anti join</b> - is opposite of Semi join - it removes records in left dataset ( tab1 ) if there are matching records in right dataset ( tab2 ).  Like semi Join, Anti Join will not bring any columns from right dataset ( tab 2 )

```{r}
anti_join(tab1, tab2, by = "state")
```

Another suported syntax

```{r}
tab1 %>% anti_join(tab2, by = "state")
```


<H3> Web Scraping </H3>

rvest package helps scape the web
```{r}
library(rvest)
```

Read the data from Wikipedia
```{r}
url <- "https://en.wikipedia.org/wiki/Murder_in_the_United_States_by_state"
h <- read_html(url)
h
```

Extract the tables from HTML code

```{r}
tabs <- html_nodes(h, "table")
tabs
```

Table the 2nd table and get the html table out of it

```{r}
tab <- tabs[[2]]
mdrs <- tab %>% html_table()
head(mdrs)
```

Lets look at the structure of the table mdrs

```{r}
str(mdrs)
```


Lets change the column names to proper names

```{r}
names(mdrs) = c("state", "population", "total_deaths", "murders", "gun_murders", "gun_ownership", "total_rate", "murder_rate", "gun_murder_rate")
str(mdrs)
```

