---
title: "String Processing"
output: html_notebook
---

This notebook is student work of String Processing module "Data Science: Wrangling" course.

<H3> Quotes and escape characters </H3>

When the string has single quote in it - we can use double quotes to define it

```{r}
s <- "He is 6' long "
cat(s)
```

When the string has double quote in it - we can use double quotes to define it

```{r}
s <- 'The scale is 10" long '
cat(s)
```


When the string has both quotes - we can need to use escape characters

```{r}
s <- 'He is 5\' 10" long '
cat(s)
```


<H3> String Processing </H3>

<B> Scape data from Web </B>

Load libraries
```{r}
library(dplyr)
library(rvest)
```

Download the web page

```{r}
url <- "https://en.wikipedia.org/wiki/Murder_in_the_United_States_by_state"
h <- read_html(url)
```

Extract the tables from HTML code

```{r}
tab <- html_nodes(h, "table")
tab <- tab[[2]]
mdrs <- tab %>% html_table()
head(mdrs)
```

Lets change the column names to proper names

```{r}
names(mdrs) = c("state", "population", "total_deaths", "murders", "gun_murders", "gun_ownership", "total_rate", "murder_rate", "gun_murder_rate")
head(mdrs)
```

Check which columns has characters

```{r}
library(stringr)
commas <- function(x) any(str_detect(x, ","))
summarise_all(mdrs, commas)
```

We know population, total_death, murders, gun_murders has , in them.  Lets replace these and convert to numeric

```{r}
mdrs$population <- as.numeric(str_replace_all(mdrs$population, ",", ""))
mdrs
```

OR use parse_number function to extract the number

```{r}
library(readr)
mdrs_new <- mdrs %>% mutate_at(2:3, parse_number)
mdrs_new
```

Now we need to clean up murders, gun_murders & murder_rate columns


<B>Handle heights data</B>

```{r}
library(dslabs)
data(reported_heights)
head(reported_heights)
```

Lets try to convert the height to numeric

```{r}
as.numeric(reported_heights$height)
```

We see a number of NA's, lets understand where these NAs are coming from

```{r}
rh <- reported_heights %>% mutate(new_height = as.numeric(height))
rh %>% filter(is.na(new_height)) %>% head(10)
```

There are many patterns above.  We will use regex to identify certain patterns and work on extracting the digital values from the input.

Lets build some examples and test out the pattern


```{r}
pve <- c("5'7\"", "4'9\"", "7' 6\"" )
nve <- c("5.7\"", "4,9\"", "3'6\"", "64" )
pattern <- "^[4-7]['][\\s]*[\\d]{1,2}[\"]*"
pattern
```

+ve test cases
```{r}
pve
```

Result
```{r}
str_detect(pve, pattern)
```

-ve test cases
```{r}
nve
```

Result
```{r}
str_detect(nve, pattern)
```
